Assessing spatial correlations in pulsar timing arrays with the noise-marginalized optimal statistic

### 2017-07-31 manuscript/manuscript.tex
more edits to Sec. III

Pulsar timing arrays (PTAs) are a conglomerate of pulsars with millisecond periods. 
Monitoring the ultra-regular periodic emission from these pulsars with radio telescopes 
enables us to probe the dynamics of the spacetime through which the pulses travel. 
PTAs are sensitive to gravitational waves (GWs) with frequencies of 
$10^{-9} - 10^{-7}\;\mathrm{Hz}$ \ki{citations needed}, 
which makes PTAs ideal complements to ground- and space-based GW detectors \sv{add citations}. 
\sv{something...} predict that the incoherent superposition of inspiraling SMBHBs form the dominant signal in this low-frequency regime.

@@

The optimal statistic is an important complement to Bayesian detection techniques. 
It provides an independent way to assess the significance of the stochastic background, 
and it is significantly faster to compute. It is also easy to test different spatial correlations 
by altering the overlap reduction function $\Gamma$ in the cross-correlations. 
In contrast, doing a full Bayesian analysis with spatial correlations is a very computationally 
intensive procedure.

@@

As introduced by \citet{abc+2009}, 
the optimal statistic is derived by analytically maximizing 
the PTA likelihood function in the weak-signal regime. 
The likelihood function is given by
\begin{equation}
	p({\bf{r}}|\phi) = \frac{\exp[-\frac{1}{2} {\bf{r}}^T \Sigma^{-1}(\phi) {\bf{r}} ]}{\sqrt{\mathrm{det}[2\pi \Sigma(\phi)]}} \,,
\end{equation}
where $\bf{r}$ is an array of the timing residuals 
from the $M$ pulsars in the PTA,
\begin{equation}
	\bf{r} = \left[ \begin{array}{c} {\bf{r}}_1 \\ {\bf{r}}_2 \\ \vdots \\ {\bf{r}}_M \end{array} \right] \,,
\end{equation}
$\phi$ is the set of pulsar noise parameters, 
and $\Sigma$ is the covariance matrix of the residuals. 
We can write $\Sigma$ in terms of 
autocovariance and cross-covariance matrices ${\bf{P}}_I$ and ${\bf{S}}_{IJ}$, respectively, as
\begin{equation}
	\Sigma = \left[ \begin{array}{cccc} {\bf{P}}_1 & {\bf{S}}_{12} & \hdots & {\bf{S}}_{1M}  \\
							{\bf{S}}_{21} & {\bf{P}}_2 & \hdots & {\bf{S}}_{2M} \\
							\vdots & \vdots & \ddots & \vdots \\
							{\bf{S}}_{M1} & {\bf{S}}_{M2} & \hdots & {\bf{P}}_M \end{array} \right] \,,
\end{equation}
where
\begin{eqnarray}
	{\bf{P}} &=& \left\langle {\bf{r}}_I {\bf{r}}_I^T \right\rangle \,, \\
	{\bf{S}}_{IJ} &=& \left. \left\langle {\bf{r}}_I {\bf{r}}_J^T \right\rangle \right|_{I \neq J} \,.
\end{eqnarray}
\sv{something about the overlap reduction function...}

@@

Figure~\ref{fig:os_dataset_sample} shows the fixed-noise and noise-marginalized 
optimal statistic for a single realization of the stochastic background, 
and Fig.~\ref{fig:os_datasetstats} compares the fixed-noise values of $\hat{A}^2$ 
to the mean $\hat{A}^2$ for each realization of the stochastic background. 
The fixed-noise analysis tends to underestimate the strength of the stochastic background, 
whereas the noise-marginalized analysis is better able to recover the injected value. 
The mean value of the optimal statistic from the fixed-noise analysis, 
averaged over realizations of the stochastic background, is $\hat{A}^2 = 8.9 \times 10^{-30}$, 
while the mean value from the noise-marginalized analysis is $\hat{A}^2 = 2.5 \times 10^{-29}$.

\begin{figure}[ht]
	\includegraphics[width=\columnwidth]{plots/os_dataset50.pdf}
	\caption{Optimal statistic and SNR for a simulated dataset 
			containing an injected GW background with $\Agw = 5\times10^{-15}$. 
			We compare the values found using a fixed-noise analysis (blue, dashed line) to the 
			mean optimal statistic and mean SNR found by marginalizing over 
			the individual pulsars' red noise parameters (orange, solid line). 
			The dotted vertical line indicates the injected value, $\hat{A}^2 = 2.5 \times 10^{-29}$. 
			For this particular realization of the stochastic background, the fixed-noise analysis 
			gives $\hat{A}^2 = 1.9 \times 10^{-30}$ with $\mathrm{SNR} = 0.60$, 
			while the noise-marginalized analysis gives $\hat{A}^2 = 1.9\times10^{-29}$ with $\mathrm{SNR} = 1.1$.}
	\label{fig:os_dataset_sample}
\end{figure}

@@

\begin{figure}[ht]
	\includegraphics[width=\columnwidth]{plots/os_datasetstats_5e-15.pdf}
	\caption{Optimal statistic and SNR for 425 simulated datasets 
			containing an injected GW background with $\Agw = 5\times10^{-15}$. 
			We compare the values found using a fixed-noise analysis (blue) to the 
			mean optimal statistic and mean SNR found by marginalizing over 
			the individual pulsars' red noise parameters. The dashed vertical line 
			indicates the injected value, $\hat{A}^2 = 2.5 \times 10^{-29}$. 
			The value of $\hat{A}^2$ from the fixed-noise analysis, 
			averaged over realizations of the stochastic background, is $\hat{A}^2 = 8.9 \times 10^{-30}$, 
			while the noise-marginalized analysis gives a mean value of $\hat{A}^2 = 2.5 \times 10^{-29}$.}
	\label{fig:os_datasetstats}
\end{figure}

@@

The optimal statistic is particularly well-suited to compare multiple spatial correlation relations 
because using a different spatial correlation only requires changing the overlap reduction function 
in $\tilde{{\bf{S}}}_{IJ}$. 
\citet{thk+2016} demonstrated how the optimal statistic can be altered to fit for 
multiple spatial correlations at once in order to mitigate common noise sources such as 
clock error and ephemeris error. 
Here we compute the optimal statistic with monopole and dipole spatial correlations 
for the same simulated data sets as in the previous section in order to determine 
how well we can distinguish a GW background from a monopole or dipole signal. 
For a monopole signal, the overlap reduction function becomes simply
\[ \Gamma(\theta_{IJ}) = 1 \,\]
while for a dipole signal, the overlap reduction function becomes
\[ \Gamma(\theta_{IJ}) = \cos\theta_{IJ} \,. \]

Figure~\ref{fig:os_ORF} shows the mean value of $\hat{A}^2$ and the mean SNR 
computed assuming monopole, dipole, and Hellings-Downs spatial correlations 
for 425 simulated data sets.

\begin{figure}[ht]
	\includegraphics[width=\columnwidth]{plots/os_datasetstats_5e-15_ORF.pdf}
	\caption{Noise-marginalized mean optimal statistic and mean SNR for 425 simulated datasets 
			containing an injected GW background with $\Agw = 5\times10^{-15}$. 
			We compare the values found using monopole (blue), dipole (orange), and quadrupole (green) 
			spatial correlations. 
			The dashed vertical line indicates the injected value, $\hat{A}^2 = 2.5 \times 10^{-29}$.}
	\label{fig:os_ORF}
\end{figure}

@@

Another way to test the spatial correlations is through ``sky scrambles,'' 
where the overlap reduction function is altered in order to simulate changing the pulsars' positions \citep{cs2016}. 
\citet{tlb+2017} showed how sky scrambles affect the Bayes' factor for simulated data sets. 
Here we perform a similar analysis using frequentist methods.

@@

\acknowledgments
We thank

JAE was supported by NASA through Einstein Fellowship Grant PF4-150120. 
SRT was supported by appointment to the NASA Postdoctoral Program 
at the Jet Propulsion Laboratory, which is administered by Oak Ridge Associated Universities 
and the Universities Space Research Association through a contract with NASA. 
KPI and SJV are supported by NSF Physics Frontier Center Grant 1430284.

>>>>

Pulsar timing arrays (PTAs) are a conglomerate of pulsars with millisecond periods. 
Monitoring the ultra-regular periodic emission from these pulsars with radio telescopes 
enables us to probe the dynamics of the spacetime through which the pulses travel. 
PTAs are sensitive to gravitational waves (GWs) with frequencies of 
$10^{-9} - 10^{-7}\;\mathrm{Hz}$ \ki{citations needed}, 
which makes PTAs ideal complements to ground- and space-based GW detectors \sv{add citations}. 
\sv{something...} predict that the incoherent superposition of inspiraling SMBHBs form the dominant signal in this low-frequency regime \sv{add citations}.

@@

The optimal statistic is an important complement to Bayesian detection techniques. 
It provides an independent way to assess the significance of the stochastic background, 
and it is significantly faster to compute. It is also easy to test different spatial correlations 
by altering the overlap reduction function (ORF) $\Gamma$ in the cross-correlations. 
In contrast, doing a full Bayesian analysis with spatial correlations is a very computationally 
intensive procedure.

@@

As introduced by \citet{abc+2009}, 
the optimal statistic is derived by analytically maximizing 
the PTA likelihood function in the weak-signal regime. 
The likelihood function is given by
\begin{equation}
	p({\bf{r}}|\phi) = \frac{\exp[-\frac{1}{2} {\bf{r}}^T \Sigma^{-1}(\phi) {\bf{r}} ]}{\sqrt{\mathrm{det}[2\pi \Sigma(\phi)]}} \,,
\end{equation}
where $\bf{r}$ is an array of the timing residuals 
from the $M$ pulsars in the PTA,
\begin{equation}
	\bf{r} = \left[ \begin{array}{c} {\bf{r}}_1 \\ {\bf{r}}_2 \\ \vdots \\ {\bf{r}}_M \end{array} \right] \,,
\end{equation}
$\phi$ is the set of pulsar noise parameters, 
and $\Sigma$ is the covariance matrix of the residuals. 
We can write $\Sigma$ in terms of 
autocovariance and cross-covariance matrices ${\bf{P}}_I$ and ${\bf{S}}_{IJ}$, respectively, as
\begin{equation}
	\Sigma = \left[ \begin{array}{cccc} {\bf{P}}_1 & {\bf{S}}_{12} & \hdots & {\bf{S}}_{1M}  \\
							{\bf{S}}_{21} & {\bf{P}}_2 & \hdots & {\bf{S}}_{2M} \\
							\vdots & \vdots & \ddots & \vdots \\
							{\bf{S}}_{M1} & {\bf{S}}_{M2} & \hdots & {\bf{P}}_M \end{array} \right] \,,
\end{equation}
where
\begin{eqnarray}
	{\bf{P}} &=& \left\langle {\bf{r}}_I {\bf{r}}_I^T \right\rangle \,, \\
	{\bf{S}}_{IJ} &=& \left. \left\langle {\bf{r}}_I {\bf{r}}_J^T \right\rangle \right|_{I \neq J} \,.
\end{eqnarray}
\sv{something about the ORF...}

@@

Figure~\ref{fig:os_dataset_sample} shows the fixed-noise and noise-marginalized 
optimal statistic for a single realization of the stochastic background, 
and Fig.~\ref{fig:os_datasetstats} compares the fixed-noise values of $\hat{A}^2$ 
to the mean $\hat{A}^2$ for each realization of the stochastic background. 
The fixed-noise analysis tends to underestimate the strength of the stochastic background, 
whereas the noise-marginalized analysis is better able to recover the injected value. 
The mean value of the optimal statistic from the fixed-noise analysis, 
averaged over realizations of the stochastic background, is $\hat{A}^2 = 8.9 \times 10^{-30}$, 
while the mean value from the noise-marginalized analysis is $\hat{A}^2 = 2.5 \times 10^{-29}$. 
The noise-marginalized analysis also gives a higher average mean SNR of 1.7, 
compared to an average SNR of 1.0 using the fixed-noise analysis.
\begin{figure}[ht]
	\includegraphics[width=\columnwidth]{plots/os_dataset50.pdf}
	\caption{Optimal statistic and SNR for a simulated dataset 
			containing an injected GW background with $\Agw = 5\times10^{-15}$. 
			We compare the values found using a fixed-noise analysis (blue, dashed line) to the 
			mean optimal statistic and mean SNR found by marginalizing over 
			the individual pulsars' red noise parameters (orange, solid line). 
			The dotted vertical line indicates the injected value, $\hat{A}^2 = 2.5 \times 10^{-29}$. 
			For this particular realization of the stochastic background, the fixed-noise analysis 
			gives $\hat{A}^2 = 1.9 \times 10^{-30}$ with $\mathrm{SNR} = 0.60$, 
			while the noise-marginalized analysis gives $\hat{A}^2 = 1.9\times10^{-29}$ with $\mathrm{SNR} = 1.1$.}
	\label{fig:os_dataset_sample}
\end{figure}

@@

\begin{figure}[ht]
	\includegraphics[width=\columnwidth]{plots/os_datasetstats_5e-15.pdf}
	\caption{Optimal statistic and SNR for 425 simulated datasets 
			containing an injected GW background with $\Agw = 5\times10^{-15}$. 
			We compare the values found using a fixed-noise analysis (blue) to the 
			mean optimal statistic and mean SNR found by marginalizing over 
			the individual pulsars' red noise parameters. The dashed vertical line 
			indicates the injected value, $\hat{A}^2 = 2.5 \times 10^{-29}$. 
			The value of $\hat{A}^2$ from the fixed-noise analysis, 
			averaged over realizations of the stochastic background, is $\hat{A}^2 = 8.9 \times 10^{-30}$, 
			while the noise-marginalized analysis gives a mean value of $\hat{A}^2 = 2.5 \times 10^{-29}$. 
			Furthermore, the fixed-noise analysis gives an average SNR of 1.0, while the 
			noise-marginalized analysis gives an average mean SNR of 1.7.}
	\label{fig:os_datasetstats}
\end{figure}

@@

The optimal statistic is particularly well-suited to compare multiple spatial correlation relations 
because using a different spatial correlation only requires changing the ORF 
in $\tilde{{\bf{S}}}_{IJ}$. 
\citet{thk+2016} demonstrated how the optimal statistic can be altered to fit for 
multiple spatial correlations at once in order to mitigate common noise sources such as 
clock error and ephemeris error. 
Here we compute the optimal statistic with monopole and dipole spatial correlations 
for the same simulated data sets as in the previous section in order to determine 
how well we can distinguish a GW background from a monopole or dipole signal. 
For a monopole signal, the ORF becomes simply
\[ \Gamma(\theta_{IJ}) = 1 \,\]
while for a dipole signal, the ORF becomes
\[ \Gamma(\theta_{IJ}) = \cos\theta_{IJ} \,. \]

Figure~\ref{fig:os_ORF} shows the noise-marginalized mean value of $\hat{A}^2$ and the mean SNR 
computed assuming monopole, dipole, and quadrupole spatial correlations 
for 425 simulated data sets. Using a monopole or dipole ORF 
gives a lower value for the mean optimal statistic and mean SNR compared to the 
Hellings-Downs ORF. 
However, there is significant overlap between the distributions for the mean SNR using 
monopole, dipole, and quadrupole ORFs, indicating that it is not easy to differentiate 
between those three spatial correlations for an injected GW background of this amplitude. 
We find a noise-marginalized mean SNR above 1.0 in 68\% of our simulated data sets 
using the quadrupole ORF, and in 45\% and 24\% of our simulated data sets 
using the monopole and dipole ORF, respectively.
\begin{figure}[ht]
	\includegraphics[width=\columnwidth]{plots/os_datasetstats_5e-15_ORF.pdf}
	\caption{Noise-marginalized mean optimal statistic and mean SNR for 425 simulated datasets 
			containing an injected GW background with $\Agw = 5\times10^{-15}$. 
			We compare the values found using monopole (blue), dipole (orange), and quadrupole (green) 
			spatial correlations. 
			The dashed vertical line indicates the injected value, $\hat{A}^2 = 2.5 \times 10^{-29}$.}
	\label{fig:os_ORF}
\end{figure}

@@

Another way to test the spatial correlations is through ``sky scrambles,'' 
where the ORF is altered in order to simulate changing the pulsars' positions \citep{cs2016}. 
\citet{tlb+2017} showed how sky scrambles affect the Bayes' factor for simulated data sets. 
Here we perform a similar analysis using frequentist methods.

@@

\acknowledgments
We thank Joe Romano and Xavier Siemens for useful discussions. 
JAE was supported by NASA through Einstein Fellowship Grant PF4-150120. 
SRT was supported by appointment to the NASA Postdoctoral Program 
at the Jet Propulsion Laboratory, which is administered by Oak Ridge Associated Universities 
and the Universities Space Research Association through a contract with NASA. 
KPI and SJV are supported by NSF Physics Frontier Center Grant 1430284.
